Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-09 E-Hypergraphs-LICS24.tex
Section 5 update

\begin{definition}[Convex DPOI rewriting]
Let $\mathfrak{R}$ be a set of DPOI rewrite rules. 
Then, given $\mathcal G \xleftarrow{} n+m$ and $\mathcal H \xleftarrow{} n + m$ in $\catname{Hyp(\Sigma)}$, $\mathcal G$ rewrites convexly into $\mathcal H$ with interface $n + m$ --- notation $(\mathcal G \xleftarrow{} n + m ) \Rrightarrow_{\mathfrak{R}}  (\mathcal H \xleftarrow{} n + m )$ --- if there exist rule $\mathcal L \xleftarrow{} i + j \xrightarrow{} \mathcal R$ in $\mathcal{R}$ and object $C$ and cospan arrows $i+j \xrightarrow{} \mathcal C \xleftarrow{} n+m$ such that the DPOI diagram of Definition \ref{def:dpoi} commutes and its marked squares are pushouts 
and the following conditions hold
\begin{itemize}
    \item $m : L \to \mathcal G$ is a convex match;
    \item $i + j \to \mathcal C \to \mathcal G$ is a boundary complement in the leftmost pushout.
\end{itemize}
\end{definition}
%By requiring convex matching and boundary complements we ensure that the original cospans above are monogamous, e.g., $n \xrightarrow{} G \xleftarrow{} m$.

@@

CHRIS: REMARK THAT WE CANNOT DEAL WITH SIGNATURES WITH $f:0 \to 0$ MORPHISMS, AND WHY!
Because extended cospans have a more general notion of interface,  we need to make some adjustments to DPOI rewriting as presented in Section \ref{sec:combinatorial-semantics}.
The first thing is that we need to account for is the existence of \textit{internal} interfaces,  and in particular that internal interfaces are not necessarily preserved during rewriting.
To illustrate this, consider the following rewrite rule
\[\langle x \xrightarrow{} x' \xrightarrow{} \mathcal{L} \xleftarrow{} y' \xleftarrow{} y, x \xrightarrow{} x'' \xrightarrow{} \mathcal{R} \xleftarrow{} y'' \xleftarrow{} y \rangle \]
such that $x' \not \cong x''$ and $y' \not \cong y''$ and consider the following e-hypergraph to perform the rewrite on
\[
    n \xrightarrow{} n' \xrightarrow{} \mathcal{G} \xleftarrow{} m' \xleftarrow{} m
\] such that there is a matching $m : \mathcal{L} \to \mathcal{G}$.
When removing the occurence of $\mathcal{L}$ from $\mathcal{G}$, e.g. by computing the pushout complement, the internal interface of $\mathcal{G}$ should be modified.  This is because the verticies of $\mathcal{L}$ corresponding to strict internal interface of $\mathcal{L}$ (and hence to strict internal interface of $\mathcal{G}$) should be removed from $\mathcal{G}$ and therefore from $\mathcal{G}$'s internal interface.
Then, when gluing $\mathcal{R}$ into the created context, we also introduce strict internal interfaces of $\mathcal{R}$ into this context.
% The input interface on the left-hand side of the rule is $\{1, 2\}$, while the input interface on the right-hand side of the rule is $\{1,2,5,6,7,8\}$.
% Note that the outermost interfaces are preserved.

@@

To account for such changes in the interfaces we require additional structure on DPOI rewriting for $\catname{EHyp({\Sigma})}$.
Consider a DPOI diagram with extra constraints below which encodes
the notion that e-hypergraph $G$ rewrites into $H$ with external interfaces $n_{ext}+m_{ext}$.
Note the addition of internal interfaces and how we use the same trick to go from cospan $n_{ext} \xrightarrow{} g_i \xrightarrow{} G \xleftarrow{} g_o \xleftarrow{} m_{ext}$ to $G \xleftarrow{} g_{i} + g_{o} \xleftarrow{} n_{ext} + m_{ext}$.
The original DPO square is depicted in blue.
% \[\begin{tikzcd}
% 	&&&& {i + j} \\
% 	{l_i+l_o} && L &&&& R && {r_i+r_o} \\
% 	\\
% 	{g_i+g_o} && {G^{\urcorner}} && C && {^{\ulcorner}H} && {h_i+h_o} \\
% 	\\
% 	&&&& {n+m}
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=2-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=2-7]
% 	\arrow["m", color={rgb,255:red,214;green,92;blue,92}, from=2-3, to=4-3]
% 	\arrow["{[i_c,j_c]}", color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=4-5]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=2-7, to=4-7]
% 	\arrow["{[n_c,m_c]}"', color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-5]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=4-5, to=4-7]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=4-5, to=4-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-7]
% 	\arrow[from=2-1, to=2-3]
% 	\arrow["{[f_i,f_o]}", from=2-1, to=4-1]
% 	\arrow[from=4-1, to=4-3]
% 	\arrow[from=2-9, to=2-7]
% 	\arrow[from=4-9, to=4-7]
% 	\arrow["{[h_i,h_o]}"{description}, from=2-9, to=4-9]
% 	\arrow[from=6-5, to=4-9]
% 	\arrow[from=6-5, to=4-1]
% 	\arrow["{[i_l,j_l]}"{description}, from=1-5, to=2-1]
% 	\arrow["{[i_r,j_r]}"{description}, from=1-5, to=2-9]
% \end{tikzcd}\]

% https://q.uiver.app/#q=WzAsMTEsWzIsMCwiTCJdLFs0LDAsInhfe2V4dH0gKyB5X3tleHR9Il0sWzIsMiwiR157XFx1cmNvcm5lcn0iXSxbNCwyLCJDIl0sWzQsNCwibl97ZXh0fSttX3tleHR9Il0sWzAsMCwibF9pK2xfbyJdLFswLDIsImdfaStnX28iXSxbNiwwLCJSIl0sWzYsMiwiXntcXHVsY29ybmVyfUgiXSxbOCwyLCJoX2kraF9vIl0sWzgsMCwicl9pK3JfbyJdLFsxLDAsIiIsMCx7ImNvbG91ciI6WzAsNjAsNjBdfV0sWzAsMiwibSIsMCx7ImNvbG91ciI6WzAsNjAsNjBdfSxbMCw2MCw2MCwxXV0sWzEsMywiW2lfYyxqX2NdIiwwLHsiY29sb3VyIjpbMCw2MCw2MF19LFswLDYwLDYwLDFdXSxbNCwzLCJbbl9jLG1fY10iLDIseyJjb2xvdXIiOlswLDYwLDYwXX0sWzAsNjAsNjAsMV1dLFszLDIsIiIsMix7ImNvbG91ciI6WzAsNjAsNjBdfV0sWzQsMiwiIiwyLHsiY29sb3VyIjpbMCw2MCw2MF19XSxbNSwwXSxbNiwyXSxbNCw4LCIiLDAseyJjb2xvdXIiOlswLDYwLDYwXX1dLFszLDgsIiIsMCx7ImNvbG91ciI6WzAsNjAsNjBdfV0sWzcsOCwiIiwyLHsiY29sb3VyIjpbMCw2MCw2MF19XSxbMSw3LCIiLDIseyJjb2xvdXIiOlswLDYwLDYwXX1dLFsxMCw3XSxbOSw4XV0=
\[
 \scalebox{0.75}{
    \tikzfig{combinatorial_semantics/DPOI_square_coloured}
 }
\]

Then, we explicitly show how the internal interfaces of the pushout complement are constructed.

@@

For monogamous extended cospans $x_{ext} \xrightarrow{} l_{i} \xrightarrow{} L \xleftarrow{} l_{o} \xleftarrow{} y_{ext}$ and $n_{ext} \xrightarrow{} g_{i} \xrightarrow{} G \xleftarrow{} g_{o} \xleftarrow{} m_{ext}$ and mono $f : L \to G$, a pushout complement $x_{ext} + y_{ext} \to C \to G^{\urcorner}$ as depicted in the square below
\[
\scalebox{0.75}{

@@

is a \textit{boundary} complement if $[i_c, i_j]$ is mono 
 and there exist $n_c : n_{ext} \to C$ and $m_c : m_{ext} \to C$ 
 making the above triangle commute and such that
there exists a \textit{not necessarily well-typed} mda-cospan

\[
n_{ext} \xrightarrow{} c_i + y_{ext} \xrightarrow{} C \xleftarrow{} c_o + x_{ext} \xleftarrow{} m_{ext}
\]

where $c_i$ and $c_o$ are computed as
 $c_i = g_i \setminus (l_i \setminus x_{ext})$ and
 $c_o = g_o \setminus (l_o \setminus y_{ext})$.

 ALEKSEI: ARE MORPHISMS $c_i + y_{ext} \to C$ AND THE SYMMETRIC ONE OBVIOUS?
\end{definition}
%
%\begin{figure}
%    \centering
%    \[
%    \scalebox{0.5}{\tikzfig{combinatorial_semantics/interfaces_change_example}}\]
%    \caption{Hypothetical DPO square for $\catname{MACsp_{D}(EHyp_{\Sigma})}$}
%    \label{fig:interface_change_example}
%\end{figure}

@@

Note that cospan $n_{ext} \xrightarrow{} c_i + y_{ext} \xrightarrow{} C \xleftarrow{} c_o + x_{ext} \xleftarrow{} m_{ext}$ is not necessarily well-typed.
The cospan's input internal interface may contain nodes that previously were a part of the output internal interface and similarly for output internal interface.

@@

\begin{definition}[Down-closed subgraph]
    We call a subgraph $\mathcal{H}$ of an e-hypergraph $\mathcal{G}$ \emph{down-closed} if for all $e \in E_{\mathcal{H}}$,   every element of $(e]$ is also in $\mathcal{H}$.
\end{definition}

\begin{definition}[Convex down-closed match] 
    We call $m : L \to G$ in $\catname{EHyp(\Sigma)}$ a convex down-closed match if

    \begin{enumerate}
        \item $m$ is monomorphism
        \item $m(L)$ is a convex down-closed subgraph of $G$
    \end{enumerate}
\end{definition}

The whole process of rewriting including the construction of the internal interfaces of the resulting e-hypergraph is summarised in the following definition.

\begin{definition}[Rewriting in $\WellTypedMdaEcospans$]
    Given $G \xleftarrow{} g_i + g_o \xleftarrow{} n_{ext} + m_{ext}$ 
    and $H \xleftarrow{} h_i + h_o \xleftarrow{} n_{ext} + m_{ext}$ in $\Ecospans$ with the outermost interfaces $n_{ext} + m_{ext}$, $G$ rewrites convexly into $H$ with the outermost interface $n_{ext} + m_{ext}$ --- notation $( G \xleftarrow{} n_{ext} + m_{ext} ) \Rrightarrow_{\mathcal{R}}^{+} (E \xleftarrow{} n_{ext} + m_{ext} )$ --- if there exists a rule $L \xleftarrow{} l_i + l_o \xleftarrow{} x_{ext} + y_{ext} \xrightarrow{} r_i + r_o \xrightarrow{} R$ with the outermost interfaces $x_{ext} + y_{ext}$ in the set of rewrite rules $\mathcal{R}$ and object $C$ and cospan arrows $x_{ext} + y_{ext} \xrightarrow{} C \xleftarrow{} n_{ext} + m_{ext}$ such that the diagram below commutes and its marked squares are pushouts

% https://q.uiver.app/#q=WzAsMTEsWzIsMCwiTCJdLFs0LDAsInhfe2V4dH0gKyB5X3tleHR9Il0sWzIsMiwiR157XFx1cmNvcm5lcn0iXSxbNCwyLCJDIl0sWzQsNCwibl97ZXh0fSttX3tleHR9Il0sWzAsMCwibF9pK2xfbyJdLFswLDIsImdfaStnX28iXSxbNiwwLCJSIl0sWzYsMiwiXntcXHVsY29ybmVyfUgiXSxbOCwyLCJoX2kraF9vIl0sWzgsMCwicl9pK3JfbyJdLFsxLDAsIiIsMCx7ImNvbG91ciI6WzAsNjAsNjBdfV0sWzAsMiwibSIsMCx7ImNvbG91ciI6WzAsNjAsNjBdfSxbMCw2MCw2MCwxXV0sWzEsMywiW2lfYyxqX2NdIiwwLHsiY29sb3VyIjpbMCw2MCw2MF19LFswLDYwLDYwLDFdXSxbNCwzLCJbbl9jLG1fY10iLDIseyJjb2xvdXIiOlswLDYwLDYwXX0sWzAsNjAsNjAsMV1dLFszLDIsIiIsMix7ImNvbG91ciI6WzAsNjAsNjBdfV0sWzQsMiwiIiwyLHsiY29sb3VyIjpbMCw2MCw2MF19XSxbNSwwXSxbNiwyXSxbNCw4LCIiLDAseyJjb2xvdXIiOlswLDYwLDYwXX1dLFszLDgsIiIsMCx7ImNvbG91ciI6WzAsNjAsNjBdfV0sWzcsOCwiIiwyLHsiY29sb3VyIjpbMCw2MCw2MF19XSxbMSw3LCIiLDIseyJjb2xvdXIiOlswLDYwLDYwXX1dLFsxMCw3XSxbOSw4XV0=
\[\scalebox{0.75}{
    \tikzfig{combinatorial_semantics/DPOI_square}
    }
\]
    and the following conditions hold

    \begin{enumerate}
        \label{dpoi:assumptions}
        \item $m : L \to G$ is a convex down-closed match;
        \item $x_{ext} + y_{ext} \to C \to G$ is a boundary complement in the leftmost pushout.
    \end{enumerate}
\end{definition}

The internal interfaces of the resulting e-hypergraph $H$ from the cospan
 $n_{ext} \xrightarrow{} h_i \xrightarrow{} H \xleftarrow{} h_o \xleftarrow{} m_{ext}$
 are constructed as $h_i = (r_i \setminus x_{ext}) + c_i$ 
 and
 $h_o = (r_o \setminus y_{ext}) + c_o$.
 $h_i \to H$ is defined as a copairing of 
 $f_1 : r_i \setminus x_{ext} \to R \to H$ 
 and 
 $f_2 : c_i \to C \to H$. 
Since both $f_1$ and $f_2$ are mono, their copairing is also a mono. 
Similarly for $h_o$.

We then say that an mda-cospan 
\[
    n_{ext} \xrightarrow{a_1} g_{i} \xrightarrow{a_2} G \xleftarrow{b_2} g_{o} \xleftarrow{b_1} m_{ext}
\]
rewrites convexly into another mda-cospan

\[
    n_{ext} \xrightarrow{a_1} h_{i} \xrightarrow{a_2} H \xleftarrow{b_2} h_{o} \xleftarrow{b_1} m_{ext}
\]

under a rewrite rule 
\[
    \langle x_{ext} \xrightarrow{a_1} l_{i} \xrightarrow{a_2} L \xleftarrow{b_2} l_{o} \xleftarrow{b_1} y_{ext}, 
    x_{ext} \xrightarrow{a_1} r_{i} \xrightarrow{a_2} R \xleftarrow{b_2} r_{o} \xleftarrow{b_1} y_{ext} \rangle
\]

if 
\begin{align*}    
    G \xleftarrow{[ a_2,b_2 ]} g_i + g_{o}& \xleftarrow{[a_1, b_1]} n_{ext} + m_{ext} \\ &\Rrightarrow^{+}\\ H \xleftarrow{[a_2,b_2]} h_i + h_{o}& \xleftarrow{[a_1, b_1]} n_{ext} + m_{ext}
\end{align*}

under a rewrite rule
\[
    \langle l_{i} + l_{o} \xrightarrow{[a_2, b_2]} L \xleftarrow{[a_1, b_1]} x_{ext} + y_{ext},
    r_{i} + r_{o} \xrightarrow{[a_2, b_2]} R \xleftarrow{[a_1, b_1]} x_{ext} + y_{ext} 
    \rangle
\]

\begin{proposition}
$n_{ext} \xrightarrow{n_h} h_i \xrightarrow{f} H \xleftarrow{g} h_o \xleftarrow{m_h} m_{ext}$ is a well-typed mda-cospan.
\end{proposition}

>>>>

\begin{definition}[Convex DPOI rewriting]
\label{def:convex_dpo}
Let $\mathfrak{R}$ be a set of DPOI rewrite rules. 
Then, given $\mathcal G \xleftarrow{} n+m$ and $\mathcal H \xleftarrow{} n + m$ in $\catname{Hyp(\Sigma)}$, $\mathcal G$ rewrites convexly into $\mathcal H$ with interface $n + m$ --- notation $(\mathcal G \xleftarrow{} n + m ) \Rrightarrow_{\mathfrak{R}}  (\mathcal H \xleftarrow{} n + m )$ --- if there exist rule $\mathcal L \xleftarrow{} i + j \xrightarrow{} \mathcal R$ in $\mathcal{R}$ and object $C$ and cospan arrows $i+j \xrightarrow{} \mathcal C \xleftarrow{} n+m$ such that the DPOI diagram of Definition \ref{def:dpoi} commutes and its marked squares are pushouts 
and the following conditions hold
\begin{itemize}
    \item $m : L \to \mathcal G$ is a convex match;
    \item $i + j \to \mathcal C \to \mathcal G$ is a boundary complement in the leftmost pushout.
\end{itemize}
\end{definition}
%By requiring convex matching and boundary complements we ensure that the original cospans above are monogamous, e.g., $n \xrightarrow{} G \xleftarrow{} m$.

@@

CHRIS: REMARK THAT WE CANNOT DEAL WITH SIGNATURES WITH $f:0 \to 0$ MORPHISMS, AND WHY!
Because extended cospans have a more general notion of interface,  we need to make some adjustments to DPOI rewriting as presented in Section \ref{sec:combinatorial-semantics}.
The first thing is that we need to account for is the existence of \textit{internal} interfaces,  and in particular that internal interfaces are not necessarily preserved during rewriting.
The former means that rewrite rules become of the form
\[
\mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j \xrightarrow{} i'' + j'' \xrightarrow{} \mathcal{R}
\] where we use the fact that cospans 
\[
    i \xrightarrow{} i' \xrightarrow{} \mathcal{L} \xleftarrow{} j' \xleftarrow{} j
\]
and
\[
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j
\]
encode the same data and we will further omit the arrows $0 \to 0 \to \mathcal{L}$.
That is, a rewrite rule is a pair of cospans of e-hypergraphs with matching \textit{external} interfaces and arbitrary \textit{internal} interfaces.
To illustrate the non-preservation of internal interfaces, consider the rewrite rule above and the following e-hypergraph with interfaces to perform the rewrite on
% \[\langle x \xrightarrow{} x' \xrightarrow{} \mathcal{L} \xleftarrow{} y' \xleftarrow{} y, x \xrightarrow{} x'' \xrightarrow{} \mathcal{R} \xleftarrow{} y'' \xleftarrow{} y \rangle \]
% such that $x' \not \cong x''$ and $y' \not \cong y''$ and consider the following e-hypergraph to perform the rewrite on
\[
    \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k
\] such that there is a matching $m : \mathcal{L} \to \mathcal{G}$  and assume $i' \not \cong i''$ and $j' \not \cong j''$.
When removing the occurence of $\mathcal{L}$ from $\mathcal{G}$, \emph{i.e.}, by computing the pushout complement, the internal interfaces of $\mathcal{G}$ should be modified to exclude the vertices that were a part of (the image of) $\mathcal{L}$'s strict internal interfaces.
Then, when gluing $\mathcal{R}$ into the created context, we need to introduce strict internal interfaces of $\mathcal{R}$ into this context.
In particular, it means that the internal interfaces of the resulting cospan of e-hypergraphs
\[
    \mathcal{H} \xleftarrow{} n'' + k'' \xleftarrow{} n + k 
\]
not necessarily coincide with the internal interfaces of the initial cospan.
% The input interface on the left-hand side of the rule is $\{1, 2\}$, while the input interface on the right-hand side of the rule is $\{1,2,5,6,7,8\}$.
% Note that the outermost interfaces are preserved.

@@

To account for such changes in the interfaces we require additional structure on DPOI rewriting for $\catname{EHyp({\Sigma})}$ which is summarised in the definition below.
\begin{definition}[Extended DPOI (EDPOI) rewriting]
\label{def:dpoi-e}
Given a span of morphisms 
\[
    \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k \xrightarrow{} n'' + k'' \xrightarrow{} \mathcal{H}
\]
in $\catname{EHyp(\Sigma)}$, we say $\mathcal{G}$ \textit{rewrites to} $\mathcal{H}$ \textit{with external interface} $n + k$ \textit{via a rewrite rule} 
\[
    \mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j \xrightarrow{} i'' + j'' \xrightarrow{} \mathcal{R}
\]
if there exists an object $\mathcal{C}$ and morphisms which complete the following commutative diagram such that the two marked squares are pushouts.
% \[\begin{tikzcd}
% 	&&&& {i + j} \\
% 	{l_i+l_o} && L &&&& R && {r_i+r_o} \\
% 	\\
% 	{g_i+g_o} && {G^{\urcorner}} && C && {^{\ulcorner}H} && {h_i+h_o} \\
% 	\\
% 	&&&& {n+m}
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=2-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=2-7]
% 	\arrow["m", color={rgb,255:red,214;green,92;blue,92}, from=2-3, to=4-3]
% 	\arrow["{[i_c,j_c]}", color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=4-5]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=2-7, to=4-7]
% 	\arrow["{[n_c,m_c]}"', color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-5]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=4-5, to=4-7]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=4-5, to=4-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-7]
% 	\arrow[from=2-1, to=2-3]
% 	\arrow["{[f_i,f_o]}", from=2-1, to=4-1]
% 	\arrow[from=4-1, to=4-3]
% 	\arrow[from=2-9, to=2-7]
% 	\arrow[from=4-9, to=4-7]
% 	\arrow["{[h_i,h_o]}"{description}, from=2-9, to=4-9]
% 	\arrow[from=6-5, to=4-9]
% 	\arrow[from=6-5, to=4-1]
% 	\arrow["{[i_l,j_l]}"{description}, from=1-5, to=2-1]
% 	\arrow["{[i_r,j_r]}"{description}, from=1-5, to=2-9]
% \end{tikzcd}\]

% https://q.uiver.app/#q=WzAsMTEsWzIsMCwiTCJdLFs0LDAsInhfe2V4dH0gKyB5X3tleHR9Il0sWzIsMiwiR157XFx1cmNvcm5lcn0iXSxbNCwyLCJDIl0sWzQsNCwibl97ZXh0fSttX3tleHR9Il0sWzAsMCwibF9pK2xfbyJdLFswLDIsImdfaStnX28iXSxbNiwwLCJSIl0sWzYsMiwiXntcXHVsY29ybmVyfUgiXSxbOCwyLCJoX2kraF9vIl0sWzgsMCwicl9pK3JfbyJdLFsxLDAsIiIsMCx7ImNvbG91ciI6WzAsNjAsNjBdfV0sWzAsMiwibSIsMCx7ImNvbG91ciI6WzAsNjAsNjBdfSxbMCw2MCw2MCwxXV0sWzEsMywiW2lfYyxqX2NdIiwwLHsiY29sb3VyIjpbMCw2MCw2MF19LFswLDYwLDYwLDFdXSxbNCwzLCJbbl9jLG1fY10iLDIseyJjb2xvdXIiOlswLDYwLDYwXX0sWzAsNjAsNjAsMV1dLFszLDIsIiIsMix7ImNvbG91ciI6WzAsNjAsNjBdfV0sWzQsMiwiIiwyLHsiY29sb3VyIjpbMCw2MCw2MF19XSxbNSwwXSxbNiwyXSxbNCw4LCIiLDAseyJjb2xvdXIiOlswLDYwLDYwXX1dLFszLDgsIiIsMCx7ImNvbG91ciI6WzAsNjAsNjBdfV0sWzcsOCwiIiwyLHsiY29sb3VyIjpbMCw2MCw2MF19XSxbMSw3LCIiLDIseyJjb2xvdXIiOlswLDYwLDYwXX1dLFsxMCw3XSxbOSw4XV0=
\[
 \scalebox{0.75}{
    \tikzfig{combinatorial_semantics/DPOI_square}
 }
\]
\end{definition}
To guarantee that the rewriting yields a monogamous directed acyclic e-hypergraph we also need to introduce a revised version of a boundary complement.

@@

For monogamous extended cospans 
\[
    i \xrightarrow{} i' \xrightarrow{} \mathcal{L} \xleftarrow{} j' \xleftarrow{} j
\] 
and 
\[
    n \xrightarrow{} n' \xrightarrow{} \mathcal{G} \xleftarrow{} k' \xleftarrow{} k
\] and mono $f : \mathcal{L} \to \mathcal{G}$, a pushout complement 
\[
    i + j \to \mathcal{L}^{\bot} \to G^{\urcorner}
\] 
as depicted in the square below
\[
\scalebox{0.75}{

@@

is a \textit{boundary} complement if $[c_1,c_2]$ is mono 
 and there exist $d_1 : n \to C$ and $d_2 : k \to C$ 
 making the above triangle commute and such that
there exists a \textit{not necessarily well-typed} mda-cospan

\[
n \xrightarrow{} l + j \xrightarrow{[g_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[g_2,c_1]} r + i \xleftarrow{} k
\]

where $l$ and $r$ are
\[
    l = n' \setminus (i' \setminus i)
\]
and
\[
    r = k' \setminus (j' \setminus j)
\]
Because $\mathcal{L}^{\bot}$ is essentially $\mathcal{G}$ with $\mathcal{L}$ removed apart from the part that has a pre-image in $i + j$ there is an obvious morphism from $l \to \mathcal{L}^{\bot}$ which is $g_1 = g_{int} |_{D}$ where
\[
    D = \{ v \in n' \text{ such that } \not \exists u \in (i' \setminus i) ~ . ~ g_{int}(v) = f'_{int};f(u) \}
\]
Similarly for $r \to \mathcal{L}^{\bot}$ which we denote as $g_2$.
Then there is a morphism $n \to l$ because there is a morphism $n \to n'$ and the external interfaces of $n'$ and $l$ are the same.
Similarly for $k \to r$.
\end{definition}

The notion of a boundary complement above is essentially the same as the one from Section~\ref{sec:combinatorial-semantics}.
The only difference is that we account for internal interfaces to require monogamous-ness and these interfaces must be computed explicitly to define the monogamous cospan.
%
%\begin{figure}
%    \centering
%    \[
%    \scalebox{0.5}{\tikzfig{combinatorial_semantics/interfaces_change_example}}\]
%    \caption{Hypothetical DPO square for $\catname{MACsp_{D}(EHyp_{\Sigma})}$}
%    \label{fig:interface_change_example}
%\end{figure}

@@

\begin{remark}
Note that cospan 
\[
    n \xrightarrow{} l + j \xrightarrow{} C \xleftarrow{} r + i \xleftarrow{} k
\] is not necessarily well-typed.
The cospan's input internal interface may contain nodes that previously were a part of the output internal interface and similarly for output internal interface.

@@

\begin{definition}[Subgraph]

A \textit{subgraph} of an e-hypergraph 
\[
    \mathcal{G} = \{V_{\mathcal{G}}, E_{\mathcal{G}}, s_{\mathcal{G}}, t_{\mathcal{G}}, l, <_{G},\consistency_{\mathcal{G}}\}
\]
is an e-hypergraph 
\[
    \mathcal{H} = \{V_{\mathcal{H}}, E_{\mathcal{H}}, s_{\mathcal{H}}, t_{\mathcal{H}}, l, <_{H},\consistency_{\mathcal{H}}\}
\] such that $V_{\mathcal{H}} \subseteq V_{\mathcal{G}}$, $E_{\mathcal{H}} \subseteq E_{\mathcal{G}}$ and such that $<_{\mathcal{H}}$ and $\consistency_{\mathcal{H}}$ both satisfy the properties from~\ref{def:e-homo}.
\end{definition}

\begin{definition}[Down-closed subgraph]
    We call a subgraph $\mathcal{H}$ of an e-hypergraph $\mathcal{G}$ \emph{down-closed} if for all $e \in E_{\mathcal{H}}$,   all children of $r$ are also in $\mathcal{H}$.
\end{definition}

\begin{definition}[Convex down-closed match] 
    We call $m : L \to G$ in $\catname{EHyp(\Sigma)}$ a \textit{convex down-closed match} if

    \begin{enumerate}
        \item $m$ is a monomorphism
        \item $m(L)$ is a convex down-closed subgraph of $G$
        \item $[m(v_i)) = [m(v_j))$ for all $v_i,v_j$ that have a pre-image in $V_{Z}$.
        \item $\consistency^{\mu}(m(v_i)) = \consistency^{\mu}(m(v_j))$ for all $v_i,v_j$ that have a pre-image in $V_{Z}$.
    \end{enumerate}
\end{definition}

The conditions (3)-(4), in particular, prohibit a match $\{f \mapsto f, g \mapsto g\}$ in $\mathcal{L} = f \otimes g$ and $\mathcal{G} = f + g$.

We are now ready to define the \textit{convex} rewriting for e-hypergraphs with interfaces which we will later use to show the equivalence between $\textsf{PROP}(\Sigma)$ and $\WellTypedMdaEcospans$.
It will be analogous to the one from Definition~\ref{def:convex_dpo} but again, to make sure the resulting cospan of e-hypergraph is monogamous we will have to construct internal interfaces separately.
We will also limit matches to convex down-closed ones both to guarantee an mda result and so that the result is a well-defined e-hypergraph.

\begin{definition}[Convex EDPOI rewriting]
    Let $\mathfrak{R}$ be a set of EDPOI rewrite rules.
    Then, given 
    \[
        \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k
    \] 
    and 
    \[
        \mathcal{H} \xleftarrow{} n'' + k'' \xleftarrow{} n + k
    \] in $\catname{EHyp(\Sigma)}$ with the external interfaces $n + k$, $\mathcal{G}$ \textit{rewrites convexly into} $\mathcal{H}$ \textit{with the outermost interface} $n + k$ --- notation $( \mathcal{G} \xleftarrow{} n + k) \Rrightarrow_{\mathfrak{R}}^{+} (\mathcal{H} \xleftarrow{} n + k)$ --- if there exists a rule 
    \[
        \mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j \xrightarrow{} i'' + j'' \xrightarrow{} \mathcal{R}
    \] 
    with the external interfaces $i + j$ in $\mathfrak{R}$ and object $\mathcal{C}$ and cospan arrows $i + j \xrightarrow{} \mathcal{C} \xleftarrow{} n + k$ 
    such that the diagram of Definition~\ref{def:dpoi-e} commutes and its marked squares are pushouts, and the following conditions hold

    \begin{enumerate}
        \label{dpoi-e:assumptions}
        \item $m : \mathcal{L} \to \mathcal{G}$ is a convex down-closed match;
        \item $i + j \to \mathcal{C} \to \mathcal{G}$ is a boundary complement in the leftmost pushout.
    \end{enumerate}
\end{definition}

% The internal interfaces of the resulting e-hypergraph $\mathcal{H}$ from the cospan
%  \[
%     \mathcal{H} \xleftarrow{} n'' + k'' \xleftarrow{} n + k
% \]
\begin{align*}
&n'' = (i'' \setminus i) + l\\
&k'' = (j'' \setminus j) + r\\
&n'' \to \mathcal{H} = [f_1,f_2] \text{ where }\\
&f_1 : i'' \setminus i \to \mathcal{R} \to \mathcal{H}\\ 
&f_2 : l \to \mathcal{C} \to \mathcal{H} 
\end{align*}
Since both $f_1$ and $f_2$ are monos, their copairing is also a mono. 
Similarly for $k'' \to \mathcal{H}$.

\begin{proposition}
    $n \xrightarrow{} n'' \xrightarrow{} \mathcal{H} \xleftarrow{} k'' \xleftarrow{} k$ is a well-typed mda-cospan.
\end{proposition}

Above we defined rewriting for cospans with empty input interfaces, when we will further build a correspondence between $\textsf{PROP}(\Sigma)^{+}$ and $\WellTypedMdaEcospans$ we will operate with cospans with non-empty input interfaces as well, so below we formulate what it means to rewrite cospans with non-empty input interfaces.

\begin{definition}[Convex EDPOI in $\WellTypedMdaEcospans$]

We say that an \textit{mda-cospan}
\[
    n \xrightarrow{} n' \xrightarrow{} \mathcal{G} \xleftarrow{} k' \xleftarrow{} k
\]
\textit{rewrites convexly into another mda-cospan}

\[
    n \xrightarrow{} n'' \xrightarrow{} \mathcal{H} \xleftarrow{} k'' \xleftarrow{} k
\]

under a rewrite rule 
% \[
%     \langle x_{ext} \xrightarrow{a_1} l_{i} \xrightarrow{a_2} L \xleftarrow{b_2} l_{o} \xleftarrow{b_1} y_{ext}, 
%     x_{ext} \xrightarrow{a_1} r_{i} \xrightarrow{a_2} R \xleftarrow{b_2} r_{o} \xleftarrow{b_1} y_{ext} \rangle
% \]

\[\mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j \xrightarrow{} i'' + j'' \xrightarrow{} \mathcal{R}\]

if 
\begin{align*}    
    \mathcal{G} \xleftarrow{} n + k \Rrightarrow^{+} \mathcal{H} \xleftarrow{} n + k
\end{align*}
under the same rewrite rule.

\begin{remark}
There is a minor limitation for EDPOI rewriting above in that it is unable to express rewriting
for rules that contain an edge with no inputs and outputs on their right-hand side (such edges correspond to morphisms of type $0 \to 0$).
This is because there is no way to impose both consistency and child relation of such an edge as it has no incident nodes in its context.
We claim that this is only a minor limitation because in practice such edges (morphisms) with no
incident nodes correspond to ‘scalars’, if we consider for example a category of vector spaces, and scalars are usually removed by being absorbed by vectors rather than introduced during rewrites.
\end{remark}
